---
title: "Stat998 - Project 2"
author: "Fred Boehm"
date: "March 12, 2015"
bibliography: stat998_2.bib
output:
  html_document:
    fig_caption:
    - yes
  pdf_document:
    fig_caption:
    - yes
---

```{r "setup", echo=FALSE}
rm(list=ls())

library(knitr)
opts_chunk$set(cache=FALSE, echo=TRUE, results= 'asis', message = TRUE, warning=TRUE, tidy=TRUE, highlight=TRUE, fig.show='asis')
```




```{r "readdat"}
library(tidyr)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
wls<- tbl_df(read.csv("WLS2.csv"))
```

```{r HA2011responders}
wls_ha2011responders <- wls[!is.na(wls$HA2011),]

```



## Using Framingham study

@d2008general use the following variables in their risk calculator (where risk is ten-year risk of cardiovascular disease event):

- age  
- hdl cholesterol  
- total cholesterol  
- Systolic Blood pressure (treated or not treated)  
- smoking  
- diabetes mellitus  

We need to identify these variables among the many variables in the Wisconsin Longitudinal Survey. 

One possible approach is to identify values of these variables at 2004, when many subjects were 65 years old (if they graduated in 1957), then ask which subjects had a 'heart attack' between 2004 and 2011. While it's only a seven-year interval, my hunch is that the established cardiovascular disease risk factors, such as those from the Framingham study [@d2008general], will be important in this study.


Q: Do we have survey questions that address the presence of other (non-MI) cardiovascular events? For instance, do we distinguish stroke from MI (myocardial infarction)? 

## References




