---
title: "Stat998 - Project 2"
author: "Fred Boehm"
date: "March 12, 2015"
output:
  pdf_document:
    fig_caption: yes
  html_document:
    fig_caption: yes
bibliography: stat998_2.bib
---

```{r "setup", echo=FALSE}
rm(list=ls())

library(knitr)
opts_chunk$set(cache=FALSE, echo=TRUE, results= 'asis', message = TRUE, warning=TRUE, tidy=TRUE, highlight=TRUE, fig.show='asis')
```




```{r "readdat"}
library(tidyr)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
wls<- tbl_df(read.csv("WLS2.csv"))
```

```{r HA2011responders}
wls_ha2011responders <- wls[!is.na(wls$HA2011),]
# find those with ha2004 & doc2004 = no & either yes or no for the 2011 ha questions
wls_nona<- wls[!is.na(select(wls, HA2011)) & !is.na(select(wls, HA2004)),]
wls_ha2004_no <- wls_nona[select(wls_nona, HA2004)=="No",]
```



## Using Framingham study

@d2008general use the following variables in their risk calculator (where risk is ten-year risk of cardiovascular disease event):

- age  
- hdl cholesterol  
- total cholesterol  
- Systolic Blood pressure (treated or not treated)  
- smoking  
- diabetes mellitus  

We need to identify these variables among the many variables in the Wisconsin Longitudinal Survey. 

One possible approach is to identify values of these variables at 2004, when many subjects were 65 years old (if they graduated in 1957), then ask which subjects had a 'heart attack' between 2004 and 2011. While it's only a seven-year interval, my hunch is that the established cardiovascular disease risk factors, such as those from the Framingham study [@d2008general], will be important in this study.


Q: Do we have survey questions that address the presence of other (non-MI) cardiovascular events? For instance, do we distinguish stroke from MI (myocardial infarction)? 

We identified those variables in the WLS that closely match those in the Framingham study [@d2008general]. It's important to note that the Framingham study used survival analysis methods, including Cox proportional hazards regression, to identify risk factors for a cardiovascular event. Thus, their study design, analysis, and purpose differ from ours. 

\begin{tabular}{ l c r }
  1 & 2 & 3 \\
  4 & 5 & 6 \\
  7 & 8 & 9 \\
\end{tabular}


\begin{table}
\begin{tabular}{l r}
Framingham study variable & WLS Variable\\
\hline
Sex & Sex\\
Quantitative total cholesterol & highchol2011 or highchol2004\\
Quantitative HDL cholesterol & None\\
Smoking & Columns 61 to 87 aim to quantify smoking \\
Diabetes & diabetes2004, diabetes2011, diabdiag2004, diabdiag2011\\
Age & Age\\
Systolic BP & highbp2004, highbp2011* \\
Treated for high blood pressure & None
\end{tabular}
\caption{Framingham study variables and their closest analogs in WLS.}
\subcaption{* SBP not available, so we used reported "high BP".}
\end{table}

## References




